---
title: "Discrete choice from data to policy, in a dozen lines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Discrete choice from data to policy, in a dozen lines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4
)
options(digits = 4)
```

**choicer** estimates discrete-choice models for applied economics. The
likelihoods, analytical gradients and Hessians are written in C++ with OpenMP,
so estimation is fast and scales to specifications with hundreds of
alternative-specific constants. Just as important, *one consistent
post-estimation interface* takes you from a fitted model to the quantities you
actually report: market shares, elasticities, diversion ratios,
willingness-to-pay with standard errors, and the welfare effects of a policy
counterfactual.

This vignette runs the whole workflow on a real data set in a few lines.

```{r setup}
library(choicer)
set_num_threads(2) # be a good CRAN citizen; raise this on your own machine
```

## The data

`mode_choice` is the classic Greene & Hensher intercity travel-mode data: 210
travellers, each choosing among **air, train, bus and car**. It ships with the
package in choicer's long layout — one row per traveller and alternative.

```{r data}
data(mode_choice)
head(mode_choice, 4)
```

`wait` (terminal waiting time) and `travel` (in-vehicle time) are in minutes;
`vcost` is the travel cost. These vary across modes within a traveller, which
is exactly what a choice model needs.

## Fit a multinomial logit

Point at the identifier, alternative and choice columns, list the covariates,
and fit. Alternative-specific constants are included by default.

```{r fit}
fit <- run_mnlogit(
  data           = mode_choice,
  id_col         = "id",
  alt_col        = "mode",
  choice_col     = "choice",
  covariate_cols = c("wait", "travel", "vcost")
)
summary(fit)
```

Estimation takes a fraction of a second. Travellers dislike waiting time,
in-vehicle time and cost — all three coefficients are negative — and the
`summary()` footer reports McFadden's R² and the in-sample hit rate alongside
the usual fit statistics.

## Predicted shares

```{r shares}
shares <- drop(predict(fit, type = "shares"))
names(shares) <- as.character(fit$alt_mapping[[2]])
round(shares, 3)

barplot(shares, col = "steelblue", ylab = "Predicted share",
        main = "Predicted mode shares")
```

## Elasticities and diversion

How does demand respond to cost, and where does it go? The same fitted object
answers both, with no extra bookkeeping.

```{r elast}
elasticities(fit, elast_var = "vcost")
diversion_ratios(fit)
```

The own-cost elasticities sit on the diagonal; off-diagonal entries are the
cross-elasticities. Because this is a plain logit, substitution follows the
independence-of-irrelevant-alternatives (IIA) property — diversion is
share-proportional. (Relaxing IIA is exactly what the [nested
logit](nl.html) and [mixed logit](mxl.html) vignettes are about.)

## Willingness to pay

With `vcost` playing the role of price, `wtp()` returns the marginal value of
each attribute in money units, with delta-method standard errors.

```{r wtp}
wtp(fit, price_var = "vcost")
```

These are the travellers' implied **value of time**: how much they would pay to
shave a minute of in-vehicle or terminal time.

## A policy counterfactual

Suppose a subsidy cuts the cost of **train** travel by 25%. Perturb the data and
predict — no refitting required.

```{r counterfactual}
mc_cf <- mode_choice
mc_cf$vcost[mc_cf$mode == "train"] <- mc_cf$vcost[mc_cf$mode == "train"] * 0.75

shares_cf <- drop(predict(fit, type = "shares", newdata = mc_cf))
names(shares_cf) <- names(shares)

rbind(baseline = shares, counterfactual = shares_cf) |> round(3)
```

Train's share rises, drawing travellers away from the other modes. We can put a
money value on the change in welfare with the expected consumer surplus:

```{r welfare}
cs0 <- consumer_surplus(fit, price_var = "vcost")
cs1 <- consumer_surplus(fit, price_var = "vcost", newdata = mc_cf)

cs1$mean_cs - cs0$mean_cs # change in mean consumer surplus
```

The cheaper train fare raises expected consumer surplus, as it should.

## One interface, every model

Everything above — `predict()`, `elasticities()`, `diversion_ratios()`,
`wtp()`, `consumer_surplus()`, plus `blp()` for share inversion — works
identically on **mixed logit** (`run_mxlogit()`) and **nested logit**
(`run_nestlogit()`) fits, and choicer also offers a Bayesian **multinomial
probit** sampler (`run_mnprobit()`). Learn each model in its own vignette:

- [Multinomial logit](mnl.html) — the workhorse, and a parameter-recovery check
- [Mixed logit](mxl.html) — random coefficients and preference heterogeneity
- [Nested logit](nl.html) — correlated alternatives and relaxed IIA
- [Bayesian multinomial probit](mnp.html) — flexible error covariance via MCMC

## How choicer compares

Several excellent R packages estimate discrete-choice models — among them
[mlogit](https://CRAN.R-project.org/package=mlogit),
[logitr](https://CRAN.R-project.org/package=logitr),
[gmnl](https://CRAN.R-project.org/package=gmnl),
[apollo](https://CRAN.R-project.org/package=apollo) and
[mixl](https://CRAN.R-project.org/package=mixl). choicer's focus is a fast C++
core with analytical gradients and Hessians, and a single, consistent
post-estimation toolkit aimed at the demand and welfare quantities applied
economists report.
