---
title: "Discrete choice from data to policy, in a dozen lines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Discrete choice from data to policy, in a dozen lines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4
)
options(digits = 4)
```

**choicer** estimates discrete-choice models for applied economics. The
likelihoods, analytical gradients and Hessians are written in C++ with OpenMP,
so estimation is fast and scales to specifications with hundreds of
alternative-specific constants. Just as important, *one consistent
post-estimation interface* takes you from a fitted model to the quantities you
actually report: market shares, elasticities, diversion ratios,
willingness-to-pay with standard errors, and the welfare effects of a policy
counterfactual.

This vignette runs the whole workflow on a real data set in a few lines.

```{r setup}
library(choicer)
set_num_threads(2) # be a good CRAN citizen; raise this on your own machine
```

## The data

`mode_choice` is the classic Greene & Hensher intercity travel-mode data: 210
travellers, each choosing among **air, train, bus and car**. It ships with the
package in choicer's long layout — one row per traveller and alternative.

```{r data}
data(mode_choice)
head(mode_choice, 4)
```

`wait` (terminal waiting time) and `travel` (in-vehicle time) are in minutes;
`vcost` is the travel cost. These vary across modes within a traveller, which
is exactly what a choice model needs.

## Fit a multinomial logit

Point at the identifier, alternative and choice columns, list the covariates,
and fit. Alternative-specific constants are included by default.

```{r fit}
fit <- run_mnlogit(
  data           = mode_choice,
  id_col         = "id",
  alt_col        = "mode",
  choice_col     = "choice",
  covariate_cols = c("wait", "travel", "vcost")
)
summary(fit)
```

Estimation takes a fraction of a second. Travellers dislike waiting time,
in-vehicle time and cost — all three coefficients are negative — and the
`summary()` footer reports McFadden's R² and the in-sample hit rate alongside
the usual fit statistics.

## Predicted shares

```{r shares}
shares <- drop(predict(fit, type = "shares"))
names(shares) <- as.character(fit$alt_mapping[[2]])
round(shares, 3)

barplot(shares, col = "steelblue", ylab = "Predicted share",
        main = "Predicted mode shares")
```

## Elasticities and diversion

How does demand respond to cost, and where does it go? The same fitted object
answers both, with no extra bookkeeping.

```{r elast}
elasticities(fit, elast_var = "vcost")
diversion_ratios(fit)
```

The own-cost elasticities sit on the diagonal; off-diagonal entries are the
cross-elasticities. In this small example every traveller faces the same set of
modes, so diversion is close to share-proportional — the textbook
independence-of-irrelevant-alternatives (IIA) pattern of logit. That proportionality
is an *individual-level* property, though, not a law of the model in aggregate:
once covariates vary across choice situations the averaged diversion patterns
move away from it (see the [MNL vignette](mnl.html#a-note-on-iia)). Where the
remaining limitations bite — substitution driven by *unobserved* tastes,
within-group correlation, or individual-level counterfactuals — the
[nested logit](nl.html), [mixed logit](mxl.html), and
[multinomial probit](mnp.html) change the source of substitution flexibility in
different ways. Those mechanisms, and their identification costs, are summarized
in [Choosing among logit models](#choosing-among-logit-models) below.

## Willingness to pay

With `vcost` playing the role of price, `wtp()` returns the marginal value of
each attribute in money units, with delta-method standard errors.

```{r wtp}
wtp(fit, price_var = "vcost")
```

These are the travellers' implied **value of time**: how much they would pay to
shave a minute of in-vehicle or terminal time.

## A policy counterfactual

Suppose a subsidy cuts the cost of **train** travel by 25%. Perturb the data and
predict — no refitting required.

```{r counterfactual}
mc_cf <- mode_choice
mc_cf$vcost[mc_cf$mode == "train"] <- mc_cf$vcost[mc_cf$mode == "train"] * 0.75

shares_cf <- drop(predict(fit, type = "shares", newdata = mc_cf))
names(shares_cf) <- names(shares)

rbind(baseline = shares, counterfactual = shares_cf) |> round(3)
```

Train's share rises, drawing travellers away from the other modes. We can put a
money value on the change in welfare with the expected consumer surplus:

```{r welfare}
cs0 <- consumer_surplus(fit, price_var = "vcost")
cs1 <- consumer_surplus(fit, price_var = "vcost", newdata = mc_cf)

cs1$mean_cs - cs0$mean_cs # change in mean consumer surplus
```

The cheaper train fare raises expected consumer surplus, as it should.
The number is a model-based demand calculation, not a causal estimate by itself:
it inherits the maintained utility specification, the price coefficient, and the
assumption that the counterfactual changes only the variables you changed in
`newdata`.

## One interface, every model

Everything above — `predict()`, `elasticities()`, `diversion_ratios()`,
`wtp()`, `consumer_surplus()`, plus `blp()` for share inversion — works
identically on **mixed logit** (`run_mxlogit()`) and **nested logit**
(`run_nestlogit()`) fits, and choicer also offers a Bayesian **multinomial
probit** sampler (`run_mnprobit()`). Learn each model in its own vignette:

- [Multinomial logit](mnl.html) — the workhorse, and a parameter-recovery check
- [Mixed logit](mxl.html) — random coefficients and preference heterogeneity
- [Nested logit](nl.html) — correlated alternatives and relaxed IIA
- [Bayesian multinomial probit](mnp.html) — flexible error covariance via MCMC

## Choosing among logit models

It is tempting to choose a model by asking "does it satisfy IIA?" — but that is
the wrong first question. IIA can be defined for any choice-probability system,
including market-wide shares. In microeconometric MNL, however, the primitive
restriction is on **individual choice probabilities conditional on covariates**:
a single decision-maker's odds between two alternatives ignore all others.
Aggregate shares need not satisfy the analogous market-level IIA property after
averaging over heterogeneous consumers or choice situations. The mixed logit, for
example, is built entirely from IIA-obeying logit kernels; its non-IIA aggregate
behavior comes from averaging those kernels over heterogeneity.

The mature econometric question is: **what variation identifies the substitution
pattern needed for the object I want to report?** A demand model should be
parsimonious and tractable, should match the substitution patterns disciplined by
the data, and — hardest of all — should deliver credible substitution in
counterfactuals you do not observe (a new product, a merger, a removed
alternative). In a counterfactual there is no data to discipline who substitutes
to what; the model's structure does all the work, which is exactly why the choice
of structure matters.

The useful way to compare the models is to ask **where each one spends its
assumptions** to generate flexible substitution:

| Model | Source of flexible substitution | Cost / where it can bite |
|---|---|---|
| **Multinomial logit** | Observed heterogeneity in covariates $X$ | Maximally disciplined by data, but rigid: cannot represent substitution from *unobserved* taste correlation, and stays share-proportional for any individual-level counterfactual. |
| **Nested logit** | Within-nest correlation in unobservables (one λ per nest) | Cheap, closed-form, globally well-behaved — but *you* must defend the nesting tree, and the model only allows correlation within the nests you draw. |
| **Mixed logit** | A distribution $f(\beta)$ of tastes (unobserved heterogeneity) | Very flexible, but the flexibility is only useful where $f$ is identified — typically through panel choices, variation in choice sets and attributes across markets (BLP), or designed experiments. |
| **Multinomial probit** | A covariance matrix for utility-difference errors | Flexible error correlation, but many covariance parameters, scale normalization, MCMC diagnostics, and substantial data demands. |

A few principles fall out of this table:

- **There is no free lunch.** Every relaxation of IIA buys flexibility by
  spending an assumption somewhere — a nesting tree, a taste distribution, an
  error covariance. You either pay with data (and accept the MNL's rigidity) or
  pay with structure (and own that structure's failure modes).

- **Start from the estimand.** Own-price elasticities, diversion ratios, WTP,
  logsum welfare, merger substitution and entry effects put different burdens on
  the model. A specification that fits observed shares can still deliver poor
  counterfactual diversion if the relevant substitution margin is supplied by
  functional form rather than empirical variation.

- **Match the model to your variation.** The mixed logit's random coefficients
  and the probit's covariance matrix are only as good as the data that identify
  them. With a thin cross-section — one choice per person, a fixed menu — prefer
  the MNL or a substantively motivated nested logit; reserve richer heterogeneity
  for settings with repeated choices, rich variation in choice sets and
  attributes across markets, or experimental designs.

- **Let the counterfactual drive the choice.** If your question hinges on *who*
  substitutes to *what* (merger simulation, a differentiated new entrant), the
  MNL's proportional substitution is a liability and the extra structure earns its
  keep. If you mainly need own-price responses and clean welfare numbers on the
  observed menu, the MNL is often the honest, robust baseline. The dangerous
  model is not always the simple one; it is the flexible one whose flexibility is
  supplied by assumptions the data cannot discipline.

Each model's own vignette develops these tradeoffs in detail:
[MNL](mnl.html#a-note-on-iia), [nested logit](nl.html),
[mixed logit](mxl.html#no-free-lunch-the-flexibility-lives-in-the-tails).

## How choicer compares

Several excellent R packages estimate discrete-choice models — among them
[mlogit](https://CRAN.R-project.org/package=mlogit),
[logitr](https://CRAN.R-project.org/package=logitr),
[gmnl](https://CRAN.R-project.org/package=gmnl),
[apollo](https://CRAN.R-project.org/package=apollo) and
[mixl](https://CRAN.R-project.org/package=mixl). choicer's focus is a fast C++
core with analytical gradients and Hessians, and a single, consistent
post-estimation toolkit aimed at the demand and welfare quantities applied
economists report.
