---
title: "Which standard errors, and when"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Which standard errors, and when}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

A demand estimate is only as reportable as its standard error. The default
maximum-likelihood variance — the inverse Hessian — is correct when the model
is correctly specified and choice situations are independent draws. Applied
work routinely violates one or both: the same decision maker contributes
several choice situations (panel and survey data), the sample was drawn by
outcome (choice-based sampling), or the likelihood is at best a useful
approximation. choicer treats the variance estimator as a choice you make —
and can revise after estimation, without refitting.

Every MNL, MXL and NL fit exposes four estimators through one interface:

| `type` | Estimator | When it is the right choice |
|---|---|---|
| `"hessian"` | Inverse analytical Hessian | Correct specification, independent choice situations. The classical default. |
| `"bhhh"` | Inverse outer product of scores (BHHH/OPG) | Same asymptotic justification as the Hessian; cheap; a common companion for simulated likelihoods. |
| `"robust"` | Huber-White sandwich $A^{-1} B A^{-1}$ | Quasi-ML under misspecification; also *the* valid variance under WESML / choice-based weighting (the meat is the weight-squared score outer product). |
| `"cluster"` | Sandwich over within-cluster sums of scores | Dependence across choice situations from the same decision maker, market, or sampling unit. |

Here the unit $i$ is the *choice situation*. If the same person supplies
several situations, cluster on the person. No small-sample correction is
applied—not a degrees-of-freedom multiplier, leverage correction, or few-cluster
reference distribution—so the number of clusters should be reasonably large.
With few clusters, the asymptotic cluster sandwich can be seriously
anti-conservative; choicer does not currently implement CR2, a wild-cluster
bootstrap, or randomization inference. Those designs require an external
small-sample procedure or a research design with a defensible higher-level
sampling argument.

```{r setup}
library(choicer)
set_num_threads(2)
```

## Dependence you should expect: panel data

The cleanest way to see what clustering buys is to build data with genuine
within-person dependence. `simulate_hmnl_data()` draws a panel in which each
person has their own taste vector $\beta_i \sim N(b, W)$ and makes several
choices with it — exactly the dependence structure a survey or scanner panel
produces.

```{r sim}
sim <- simulate_hmnl_data(N = 500, T = 8, J = 4, seed = 3)
```

We deliberately fit a plain multinomial logit. Its point estimates target the
population-average taste, but its likelihood treats all `500 * 8 = 4000`
choice situations as independent — which they are not, because each person's
taste deviation persists across their eight tasks. Passing `cluster_col`
supplies the person labels at fit time and selects cluster-robust standard
errors for `summary()`:

```{r fit}
fit <- run_mnlogit(
  data                   = sim$data,
  id_col                 = "task",
  alt_col                = "alt",
  choice_col             = "choice",
  covariate_cols         = c("x1", "x2"),
  cluster_col            = "pid",
  include_outside_option = TRUE
)
summary(fit)
```

How much did it matter? `vcov(fit, type = )` recomputes any of the four
estimators from the stored per-situation scores:

```{r compare}
se_of <- function(V) sqrt(diag(V))
tab <- cbind(
  hessian = se_of(vcov(fit, type = "hessian")),
  bhhh    = se_of(vcov(fit, type = "bhhh")),
  robust  = se_of(vcov(fit, type = "robust")),
  cluster = se_of(vcov(fit, type = "cluster"))
)
round(tab, 4)
round(tab[, "cluster"] / tab[, "hessian"], 2)
```

The clustered standard errors on the slope coefficients are about 25% wider
than the Hessian ones; the alternative-specific constants barely move. That
pattern is the fingerprint of the data-generating process: the persistent
component is each person's *taste* deviation, so it is the slope scores that
are correlated within person. Note also that `"robust"` alone recovers almost
none of this — the Huber-White sandwich guards against misspecification of
each situation's likelihood, not against dependence *across* situations. If
the same decision maker appears repeatedly, robust-without-clustering is the
wrong comfort.

## Post hoc, without refitting

Nothing above required deciding at fit time. Any fit stored with
`keep_data = TRUE` (the default) can be re-inferenced later — useful when the
clustering level is itself a robustness question:

```{r posthoc}
fit0 <- run_mnlogit(
  data                   = sim$data,
  id_col                 = "task",
  alt_col                = "alt",
  choice_col             = "choice",
  covariate_cols         = c("x1", "x2"),
  include_outside_option = TRUE
)

# one label per choice situation, in the prepared (id-sorted) order
cl <- sim$data$pid[!duplicated(sim$data$task)]
se_of(vcov(fit0, type = "cluster", cluster = cl))
```

The `cluster` vector has one entry per choice situation, aligned with the
situations sorted by the id column — here, one `pid` per `task`.

## Clustering repairs inference, not the estimand

A caveat that matters for the mixed logit. `run_mxlogit()` maximizes a
*cross-sectional* simulated likelihood: each choice situation is treated as an
independent draw from the mixing distribution. On panel data,
`type = "cluster"` makes the standard errors robust to within-person
dependence, but the point estimates still target that cross-sectional model —
clustering does not turn the fit into a panel mixed logit, in which one taste
draw is shared across a person's choices. When the panel structure is the
object of interest — individual-level tastes, their distribution, persistence
— estimate the panel model instead: the hierarchical Bayesian logit
(`run_hmnlogit()` with `person_col`), covered in the
[hierarchical Bayes vignette](hb.html).

## Choice-based sampling

When the sample was drawn by outcome (interviewing travellers at the terminal
of the mode they chose, oversampling rare hospitals), the weights are not
optional and the variance choice is forced: under WESML weighting the inverse
weighted Hessian and the ordinary BHHH variance are both invalid, and
`type = "robust"` — whose meat carries the squared weights — is the correct
estimator. The [choice-based sampling vignette](wesml.html) develops the full
workflow, from population shares to weighted fit.

## References

Berndt, E., Hall, B., Hall, R. and Hausman, J. (1974). Estimation and
inference in nonlinear structural models. *Annals of Economic and Social
Measurement*, 3(4), 653-665.

Cameron, A. C. and Miller, D. L. (2015). A practitioner's guide to
cluster-robust inference. *Journal of Human Resources*, 50(2), 317-372.

Manski, C. F. and Lerman, S. R. (1977). The estimation of choice
probabilities from choice based samples. *Econometrica*, 45(8), 1977-1988.

White, H. (1982). Maximum likelihood estimation of misspecified models.
*Econometrica*, 50(1), 1-25.
