---
title: "Which standard errors, and when"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Which standard errors, and when}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

A demand estimate is only as reportable as its standard error. The default
maximum-likelihood variance — the inverse Hessian — is correct when the model
is correctly specified and choice situations are independent draws. Applied
work routinely violates one or both: the same decision maker contributes
several choice situations (panel and survey data), the sample was drawn by
outcome (choice-based sampling), or the likelihood is at best a useful
approximation. choicer treats the variance estimator as a choice you make —
and can revise after estimation, without refitting.

Every MNL, MXL and NL fit exposes four estimators through one interface:

| `type` | Estimator | When it is the right choice |
|---|---|---|
| `"hessian"` | Inverse analytical Hessian | Correct specification, independent choice situations. The classical default. |
| `"bhhh"` | Inverse outer product of scores (BHHH/OPG) | Same asymptotic justification as the Hessian; cheap; a common companion for simulated likelihoods. |
| `"robust"` | Huber-White sandwich $A^{-1} B A^{-1}$ | Quasi-ML under misspecification; also *the* valid variance under WESML / choice-based weighting (the meat is the weight-squared score outer product). |
| `"cluster"` | Sandwich over within-cluster sums of scores | Dependence across choice situations from the same decision maker, market, or sampling unit. |

Here the unit $i$ is the *choice situation* — or, for a mixed logit fitted
with `person_col` (see below), the *decision maker*: several situations
already collapse to one scored unit, so there is nothing left within a person
to cluster on. Otherwise, if the same person supplies several situations,
cluster on the person. No small-sample correction is
applied—not a degrees-of-freedom multiplier, leverage correction, or few-cluster
reference distribution—so the number of clusters should be reasonably large.
With few clusters, the asymptotic cluster sandwich can be seriously
anti-conservative; choicer does not currently implement CR2, a wild-cluster
bootstrap, or randomization inference. Those designs require an external
small-sample procedure or a research design with a defensible higher-level
sampling argument.

```{r setup}
library(choicer)
set_num_threads(2)
```

## Dependence you should expect: panel data

The cleanest way to see what clustering buys is to build data with genuine
within-person dependence. `simulate_hmnl_data()` draws a panel in which each
person has their own taste vector $\beta_i \sim N(b, W)$ and makes several
choices with it — a stylized version of the persistent within-person dependence
in a survey or scanner panel.

```{r sim}
sim <- simulate_hmnl_data(N = 500, T = 8, J = 4, seed = 3)
```

We deliberately fit a plain multinomial logit. Because a homogeneous logit is
misspecified for this random-coefficients DGP, its point estimates target the
pseudo-true homogeneous-logit projection, not generally the population mean
$b$. Its likelihood also treats all `500 * 8 = 4000` choice situations as
independent — which they are not, because each person's taste deviation persists
across their eight tasks. Passing `cluster_col`
supplies the person labels at fit time and selects cluster-robust standard
errors for `summary()`:

```{r fit}
fit <- run_mnlogit(
  data                   = sim$data,
  id_col                 = "task",
  alt_col                = "alt",
  choice_col             = "choice",
  covariate_cols         = c("x1", "x2"),
  cluster_col            = "pid",
  include_outside_option = TRUE
)
summary(fit)
```

How much did it matter? `vcov(fit, type = )` recomputes any of the four
estimators from the data stored with the fit:

```{r compare}
se_of <- function(V) sqrt(diag(V))
tab <- cbind(
  hessian = se_of(vcov(fit, type = "hessian")),
  bhhh    = se_of(vcov(fit, type = "bhhh")),
  robust  = se_of(vcov(fit, type = "robust")),
  cluster = se_of(vcov(fit, type = "cluster"))
)
round(tab, 4)
round(tab[, "cluster"] / tab[, "hessian"], 2)
```

The clustered standard errors on the slope coefficients are about 25% wider
than the Hessian ones; those on the alternative-specific constants differ
little. That
pattern is the fingerprint of the data-generating process: the persistent
component is each person's *taste* deviation, so it is the slope scores that
are correlated within person. Note also that `"robust"` alone recovers almost
none of this — the Huber-White sandwich guards against misspecification of
each situation's likelihood, not against dependence *across* situations. If
the same decision maker appears repeatedly, robust-without-clustering is the
wrong comfort (the exception is a mixed logit fitted with `person_col`: there
the decision maker *is* the likelihood unit, so `"robust"` is already
clustered by construction — see below).

## Post hoc, without refitting

Nothing above required deciding at fit time. Any fit stored with
`keep_data = TRUE` (the default) can be re-inferenced later — useful when the
clustering level is itself a robustness question:

```{r posthoc}
fit0 <- run_mnlogit(
  data                   = sim$data,
  id_col                 = "task",
  alt_col                = "alt",
  choice_col             = "choice",
  covariate_cols         = c("x1", "x2"),
  include_outside_option = TRUE
)

# one cluster label per choice situation, named by choice-situation id
task_person <- unique(sim$data[, c("task", "pid")])
cl <- setNames(task_person$pid, task_person$task)
se_of(vcov(fit0, type = "cluster", cluster = cl))
```

The `cluster` vector has one entry per choice situation — here, one `pid` per
`task` — and is named by the choice-situation id. choicer uses those names to
realign labels safely to its prepared order.

## Clustering repairs inference, not efficiency

A caveat that matters for the mixed logit. Without `person_col`,
`run_mxlogit()` maximizes a *cross-sectional* simulated likelihood: each
choice situation's probability is integrated over the mixing distribution
separately, rather than holding one simulation draw fixed across a person's
tasks and integrating their joint probability. On panel data this is a
composite (pseudo-) likelihood: every situation's marginal probability is
still correctly specified, so the point estimates stay consistent for the
population mixing distribution, but the fit is statistically inefficient — it
never uses the fact that the same tastes generated a person's repeated
choices, which is exactly the within-person variation that disciplines
$\Sigma$. `type = "cluster"` (or `cluster_col` at fit time) makes the
standard errors valid under that within-person dependence, but leaves the
point estimates, and their inefficiency, unchanged: clustering repairs the
*inference*, not the *likelihood*. And because it is a different likelihood,
its log-likelihood, AIC and BIC are not comparable to a panel fit's.

When the panel structure is the object of interest — a taste distribution
disciplined by repeated choices, not just valid standard errors around a
composite fit — estimate the panel likelihood instead. Two options hold one
taste draw fixed within each person (Revelt and Train 1998):
`run_mxlogit(..., person_col = )`, covered in the
[mixed logit vignette](mxl.html), whose scores are then already per decision
maker, so `type = "robust"` is clustered by construction; or the hierarchical
Bayesian logit, `run_hmnlogit(..., person_col = )`, covered in the
[hierarchical Bayes vignette](hb.html), which adds partial pooling of
alternative-level effects and posterior rather than asymptotic uncertainty.

## Choice-based sampling

When the sample was drawn by outcome (interviewing travellers at the terminal
of the mode they chose, oversampling rare hospitals), the weights are not
optional and the variance choice is forced: under WESML weighting the inverse
weighted Hessian and the ordinary BHHH variance are both invalid, and
`type = "robust"` — whose meat carries the squared weights — is the correct
estimator, conditional on the supplied population shares `Q`. If weighted
observations are also dependent within sampling units, use the weighted cluster
sandwich instead; if `Q` is estimated, its uncertainty requires an additional
resampling or sensitivity layer. The [choice-based sampling vignette](wesml.html)
develops the full workflow, from population shares to weighted fit.

## References

Berndt, E., Hall, B., Hall, R. and Hausman, J. (1974). Estimation and
inference in nonlinear structural models. *Annals of Economic and Social
Measurement*, 3(4), 653-665.

Cameron, A. C. and Miller, D. L. (2015). A practitioner's guide to
cluster-robust inference. *Journal of Human Resources*, 50(2), 317-372.

Manski, C. F. and Lerman, S. R. (1977). The estimation of choice
probabilities from choice based samples. *Econometrica*, 45(8), 1977-1988.

Revelt, D. and Train, K. (1998). Mixed logit with repeated choices:
households' choices of appliance efficiency level. *Review of Economics and
Statistics*, 80(4), 647-657.

White, H. (1982). Maximum likelihood estimation of misspecified models.
*Econometrica*, 50(1), 1-25.
