---
title: "Multinomial logit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multinomial logit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

The multinomial logit (MNL) is the workhorse of discrete choice. choicer fits it
by maximum likelihood with a C++ core, analytical gradients and an analytical
Hessian, so estimation and standard errors are fast even with many
alternative-specific constants.

This vignette does two things: it shows the full MNL workflow, and — because the
data are simulated from a known process — it checks that choicer **recovers the
true parameters**.

```{r setup}
library(choicer)
set_num_threads(2)
```

## Simulate a known data-generating process

`simulate_mnl_data()` draws choices from a logit model with i.i.d. Gumbel
errors. The returned object carries both the data and the true parameters.

```{r sim}
sim <- simulate_mnl_data(N = 2000, J = 4, seed = 1)
sim
```

## Fit

```{r fit}
fit <- run_mnlogit(
  data           = sim$data,
  id_col         = "id",
  alt_col        = "alt",
  choice_col     = "choice",
  covariate_cols = c("x1", "x2")
)
summary(fit)
```

## Did we recover the truth?

`recovery_table()` lines up each estimate against the value that generated the
data, with the bias, a z-score, and whether the 95% confidence interval covers
the truth.

```{r recovery}
recovery_table(fit, sim$true_params)
```

In this run the intervals cover their true values, which is indicative of
correct behavior. What matters formally, though, is coverage over repeated
simulations; a full Monte Carlo exercise is left outside this vignette for
brevity.

## Post-estimation

The full demand and welfare toolkit is available on the fitted object. Treating
`x2` as price:

```{r post}
predict(fit, type = "shares")        # predicted market shares
elasticities(fit, elast_var = "x2")  # own- and cross-price elasticities
diversion_ratios(fit)                # where demand goes
wtp(fit, price_var = "x2")           # willingness to pay, with delta-method SEs
gof(fit)                             # McFadden R2 and hit rate
```

## A note on IIA

It is worth being precise about what IIA is and where it bites, because the
usual one-liner ("logit assumes IIA, so it's wrong") obscures more than it
explains.

**IIA is a property of the *individual* choice probabilities.** For a single
decision-maker, the ratio of any two choice probabilities,
$P_{ij}/P_{ik} = \exp(V_{ij} - V_{ik})$, depends only on the attributes of $j$
and $k$ — never on what other alternatives exist or what they look like. That is
the independence-of-irrelevant-alternatives axiom, and the MNL satisfies it by
construction.

**At the aggregate level the picture is different.** The diversion ratios and
cross-elasticities reported above are sample averages over individuals. The
familiar share-proportional formula $DR(j\to k) = s_k / (1 - s_j)$ holds *only*
when every individual faces the same covariates, so that $P_{ij}$ does not vary
across $i$. As soon as covariates vary across choice situations — different
attributes, demographics, or choice sets — the aggregate substitution patterns
are **no longer share-proportional**, even though each individual still obeys
IIA. (The diversion derivation in the
[math companion](multinomial_logit_math.html) makes this explicit.)

So the MNL is more flexible in aggregate than "IIA" suggests — but only along
dimensions you have *observed*. Its genuine limitations are two. First, all of
its substitution flexibility must be carried by observed covariates; it cannot
represent substitution driven by *unobserved* correlation in tastes. Second, for
a single decision-maker — and therefore for any individual-level
counterfactual — substitution remains rigidly proportional. That second point
matters most for counterfactuals (a new product, a removed alternative), where
there is no data to discipline the substitution pattern and the model's
structure does all the work.

When those limitations bite — close substitutes, rich unobserved heterogeneity,
counterfactuals that hinge on *who* substitutes to *what* — reach for the
[nested logit](nl.html) or [mixed logit](mxl.html). They keep the same
`predict()` / `elasticities()` / `wtp()` interface, and the
[getting-started vignette](choicer.html#choosing-among-logit-models) lays out
what each relaxation costs.
