---
title: "Multinomial logit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multinomial logit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

The multinomial logit (MNL) is the workhorse of discrete choice. choicer fits it
by maximum likelihood with a C++ core, analytical gradients and an analytical
Hessian, so estimation and standard errors are fast even with many
alternative-specific constants.

This vignette does two things: it shows the full MNL workflow, and — because the
data are simulated from a known process — it checks that choicer **recovers the
true parameters**.

```{r setup}
library(choicer)
set_num_threads(2)
```

## Simulate a known data-generating process

`simulate_mnl_data()` draws choices from a logit model with i.i.d. Gumbel
errors. The returned object carries both the data and the true parameters.

```{r sim}
sim <- simulate_mnl_data(N = 2000, J = 4, seed = 1)
sim
```

## Fit

```{r fit}
fit <- run_mnlogit(
  data           = sim$data,
  id_col         = "id",
  alt_col        = "alt",
  choice_col     = "choice",
  covariate_cols = c("x1", "x2")
)
summary(fit)
```

## Did we recover the truth?

`recovery_table()` lines up each estimate against the value that generated the
data, with the bias, a z-score, and whether the 95% confidence interval covers
the truth.

```{r recovery}
recovery_table(fit, sim$true_params)
```

Every interval should cover its true value — the estimator is doing its job.

## Post-estimation

The full demand and welfare toolkit is available on the fitted object. Treating
`x2` as price:

```{r post}
predict(fit, type = "shares")        # predicted market shares
elasticities(fit, elast_var = "x2")  # own- and cross-price elasticities
diversion_ratios(fit)                # where demand goes
wtp(fit, price_var = "x2")           # willingness to pay, with delta-method SEs
gof(fit)                             # McFadden R2 and hit rate
```

## A note on IIA

In the elasticity and diversion output above, substitution is
share-proportional: this is the independence-of-irrelevant-alternatives (IIA)
property of logit. It is a strong assumption. When it is implausible — close
substitutes, or heterogeneous tastes — reach for the
[nested logit](nl.html) or [mixed logit](mxl.html), which relax IIA while
keeping the same `predict()` / `elasticities()` / `wtp()` interface.
