---
title: "Bayesian multinomial probit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian multinomial probit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5, fig.height = 4)
options(digits = 4)
```

The multinomial probit (MNP) replaces logit's Type-I extreme-value error
structure with a multivariate-normal covariance for utility differences. That
covariance can encode correlated substitution across alternatives. The cost is
scale normalization, a covariance parameterization that grows quickly with the
number of alternatives, and MCMC diagnostics. choicer estimates the model the
Bayesian way: a Gibbs sampler with data augmentation, written in C++ with a
reproducible, thread-safe random number generator.

```{r setup}
library(choicer)
set_num_threads(2)
```

## Simulate from a probit process

`simulate_mnp_data()` draws choices with correlated normal errors and known
parameters.

```{r sim}
sim <- simulate_mnp_data(N = 2000, J = 3, seed = 1)
sim
```

## Run the sampler

`run_mnprobit()` returns posterior draws. The settings below keep this vignette
quick; for real work use a longer run, repeat the fit with several independent
seeds when needed, and inspect trace, autocorrelation, ESS and Monte Carlo error
before interpreting posterior summaries. `run_mnprobit()` does not currently
expose user-specified starts or a multi-chain wrapper.

```{r fit}
set.seed(3)
fit <- run_mnprobit(
  data           = sim$data,
  id_col         = "id",
  alt_col        = "alt",
  choice_col     = "choice",
  covariate_cols = c("x1", "x2"),
  mcmc           = list(R = 4000, burn = 1000, thin = 2)
)
summary(fit)
```

The `Sigma` entries are the error-covariance parameters. They are identified only
up to scale, so choicer reports them on the normalized scale where
`Sigma_11 = 1`. The prior is placed on the unrestricted covariance used inside
the Gibbs sampler; the reported posterior summaries are computed after
normalizing each kept draw.

For empirical work, ask what disciplines the covariance before interpreting
it. Keane (1992) shows that when covariates do not vary across alternatives —
demographics, or constants alone — the covariance parameters are identified by
the normal functional form only, and estimation is fragile. Alternative-specific
covariate variation (prices, distances, or attributes that differ across the
options within a choice situation), and especially defensible exclusions across
utility equations, is what separates covariance from systematic utility in
practice; the simulated data above contain two varying attributes. The Bayesian
machinery
does not repeal this: with weak alternative-specific variation the $\Sigma$
posterior sits close to its prior, and the substitution pattern it implies is
maintained rather than estimated. The MNP is attractive when flexible error
correlation is the central object, but with many alternatives its parameter
count rises quickly.

## Compare posterior summaries with the truth

```{r recovery}
recovery_table(fit, sim$true_params)
```

In this simulated example the posterior summaries line up with the parameters
that generated the data. As with the frequentist recovery vignettes, this is an
illustration rather than a proof: what matters in repeated use is posterior
coverage, mixing, and sensitivity to prior and normalization choices.

## A quick MCMC diagnostic

Posterior draws are stored on the fit, so the usual diagnostics are a line away.
A trace plot of the price coefficient should look like a stationary "fuzzy
caterpillar":

```{r trace}
beta_draws <- fit$draws$beta
plot(beta_draws[, "x2"], type = "l", col = "steelblue",
     xlab = "iteration", ylab = expression(beta[x2]),
     main = "Posterior trace: price coefficient")
abline(h = sim$true_params$beta[2], col = "red", lwd = 2)
```

The red line marks the true value; the chain should hover around it. The same
diagnostics used for the hierarchical models apply to any draws matrix:
`rhat()` computes the split R-hat (values near 1 are consistent with
convergence) and `ess()` the rank-normalized bulk and tail effective sample
sizes — the number of effectively independent draws behind each posterior
summary:

```{r rhat}
rhat(fit$draws$beta)
ess(fit$draws$beta)
mcse(fit$draws$beta)
```

On a single chain these are necessary rather than sufficient checks; the
stronger practical check in the current API is to repeat the complete fit with
several independent seeds, then compare posterior summaries and pass equal-length
draw matrices as a list to the same diagnostic functions, for example
`rhat(list(fit$draws$beta, fit_seed_2$draws$beta), rank = TRUE)`. Because the
sampler's
initial state is internal rather than user-controlled, this is not equivalent
to deliberately overdispersed starting values. Longer runs, posterior stability
across seeds, and prior sensitivity should therefore be reported together.

## What this fit does not provide

`choicer_mnp` is a posterior-draws object, not a `choicer_fit`. It provides
coefficient and covariance summaries, recovery tools, and access to the raw and
identified draws. It does **not** currently implement `predict()`, elasticities,
diversion ratios, logsum welfare, `logLik()`, AIC, or BIC. General MNP
counterfactual prediction requires multivariate-normal probability evaluation
and an explicit rule for transporting the covariance structure to a new choice
set; choicer does not silently impose either.

For a reportable MNP application, document the base alternative and per-draw
scale normalization, the prior on the raw covariance and its implications after
normalization, the alternative-specific variation that identifies covariance,
trace/ESS/MCSE evidence, stability across independent seeds, and sensitivity to
the prior and base alternative. Covariance conclusions are often the first to
move when the identifying variation is weak.

The full derivations — the data-augmented Gibbs sampler, the McCulloch-Rossi
normalization, and the identification discussion — are in the
[math companion](bayesian_multinomial_probit_math.html). For a faster,
frequentist alternative with the same post-estimation toolkit, see the
[multinomial logit](mnl.html) and [mixed logit](mxl.html) vignettes. For the
hierarchical extension — respondent-level random tastes and partially-pooled
alternative effects, in both logit and probit flavors — see the
[hierarchical Bayes vignette](hb.html).

## References

Albert, J. H. and Chib, S. (1993). Bayesian analysis of binary and
polychotomous response data. *Journal of the American Statistical
Association*, 88(422), 669-679.

Keane, M. P. (1992). A note on identification in the multinomial probit
model. *Journal of Business & Economic Statistics*, 10(2), 193-200.

McCulloch, R. and Rossi, P. E. (1994). An exact likelihood analysis of the
multinomial probit model. *Journal of Econometrics*, 64(1-2), 207-240.

Rossi, P. E., Allenby, G. M. and McCulloch, R. (2005). *Bayesian Statistics
and Marketing*. Wiley.

Train, K. E. (2009). *Discrete Choice Methods with Simulation* (2nd ed.).
Cambridge University Press, Chapter 5.
