---
title: "Bayesian multinomial probit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian multinomial probit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5, fig.height = 4)
options(digits = 4)
```

The multinomial probit (MNP) replaces logit's restrictive error structure with a
*full multivariate-normal* covariance, so it can capture rich correlation
between alternatives without imposing IIA. choicer estimates it the Bayesian way:
a Gibbs sampler with data augmentation, written in C++ with a reproducible,
thread-safe random number generator.

```{r setup}
library(choicer)
set_num_threads(2)
```

## Simulate from a probit process

`simulate_mnp_data()` draws choices with correlated normal errors and known
parameters.

```{r sim}
sim <- simulate_mnp_data(N = 2000, J = 3, seed = 1)
sim
```

## Run the sampler

`run_mnprobit()` returns posterior draws. The settings below keep this vignette
quick; for real work use longer chains, multiple starting values, and convergence
diagnostics before interpreting posterior summaries.

```{r fit}
set.seed(3)
fit <- run_mnprobit(
  data           = sim$data,
  id_col         = "id",
  alt_col        = "alt",
  choice_col     = "choice",
  covariate_cols = c("x1", "x2"),
  mcmc           = list(R = 4000, burn = 1000, thin = 2)
)
summary(fit)
```

The `Sigma` entries are the error-covariance parameters — the flexibility that
distinguishes probit from logit. They are identified only up to scale, so
choicer reports them on the normalized scale where `Sigma_11 = 1`. The prior is
placed on the unrestricted covariance used inside the Gibbs sampler; the reported
posterior summaries are computed after normalizing each kept draw.

## Compare posterior summaries with the truth

```{r recovery}
recovery_table(fit, sim$true_params)
```

In this simulated example the posterior summaries line up with the parameters
that generated the data. As with the frequentist recovery vignettes, this is an
illustration rather than a proof: what matters in repeated use is posterior
coverage, mixing, and sensitivity to prior and normalization choices.

## A quick MCMC diagnostic

Posterior draws are stored on the fit, so the usual diagnostics are a line away.
A trace plot of the price coefficient should look like a stationary "fuzzy
caterpillar":

```{r trace}
beta_draws <- fit$draws$beta
plot(beta_draws[, "x2"], type = "l", col = "steelblue",
     xlab = "iteration", ylab = expression(beta[x2]),
     main = "Posterior trace: price coefficient")
abline(h = sim$true_params$beta[2], col = "red", lwd = 2)
```

The red line marks the true value; the chain should hover around it. For a
faster, frequentist alternative with the same post-estimation toolkit, see the
[multinomial logit](mnl.html) and [mixed logit](mxl.html) vignettes.
