---
title: "Bayesian multinomial probit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian multinomial probit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5, fig.height = 4)
options(digits = 4)
```

The multinomial probit (MNP) replaces logit's Type-I extreme-value error
structure with a multivariate-normal covariance for utility differences. That
covariance can encode correlated substitution across alternatives. The cost is
scale normalization, a covariance parameterization that grows quickly with the
number of alternatives, and MCMC diagnostics. choicer estimates the model the
Bayesian way: a Gibbs sampler with data augmentation, written in C++ with a
reproducible, thread-safe random number generator.

```{r setup}
library(choicer)
set_num_threads(2)
```

## Simulate from a probit process

`simulate_mnp_data()` draws choices with correlated normal errors and known
parameters.

```{r sim}
sim <- simulate_mnp_data(N = 2000, J = 3, seed = 1)
sim
```

## Run the sampler

`run_mnprobit()` returns posterior draws. The settings below keep this vignette
quick; for real work use longer chains, multiple starting values, and convergence
diagnostics before interpreting posterior summaries.

```{r fit}
set.seed(3)
fit <- run_mnprobit(
  data           = sim$data,
  id_col         = "id",
  alt_col        = "alt",
  choice_col     = "choice",
  covariate_cols = c("x1", "x2"),
  mcmc           = list(R = 4000, burn = 1000, thin = 2)
)
summary(fit)
```

The `Sigma` entries are the error-covariance parameters. They are identified only
up to scale, so choicer reports them on the normalized scale where
`Sigma_11 = 1`. The prior is placed on the unrestricted covariance used inside
the Gibbs sampler; the reported posterior summaries are computed after
normalizing each kept draw.

For empirical work, the covariance matrix should be read as a maintained
substitution structure unless the data contain enough variation to discipline it.
The MNP is attractive when flexible error correlation is the central object, but
with many alternatives its parameter count rises quickly.

## Compare posterior summaries with the truth

```{r recovery}
recovery_table(fit, sim$true_params)
```

In this simulated example the posterior summaries line up with the parameters
that generated the data. As with the frequentist recovery vignettes, this is an
illustration rather than a proof: what matters in repeated use is posterior
coverage, mixing, and sensitivity to prior and normalization choices.

## A quick MCMC diagnostic

Posterior draws are stored on the fit, so the usual diagnostics are a line away.
A trace plot of the price coefficient should look like a stationary "fuzzy
caterpillar":

```{r trace}
beta_draws <- fit$draws$beta
plot(beta_draws[, "x2"], type = "l", col = "steelblue",
     xlab = "iteration", ylab = expression(beta[x2]),
     main = "Posterior trace: price coefficient")
abline(h = sim$true_params$beta[2], col = "red", lwd = 2)
```

The red line marks the true value; the chain should hover around it. For a
faster, frequentist alternative with the same post-estimation toolkit, see the
[multinomial logit](mnl.html) and [mixed logit](mxl.html) vignettes.
