---
title: "Mixed logit and preference heterogeneity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed logit and preference heterogeneity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

The mixed (random-coefficients) logit lets tastes vary across people. Instead of
a single coefficient on each random attribute, choicer estimates a *distribution*
of coefficients, which relaxes IIA and allows realistic substitution
patterns. Estimation is by simulated maximum likelihood using Halton draws, with
the likelihood, gradient and Hessian evaluated in parallel C++.

A useful way to see *how* it relaxes IIA: the mixed-logit probability is a
logit kernel averaged over the taste distribution,
$P_{ij} = \int L_{ij}(\beta)\, f(\beta)\, d\beta$, where
$L_{ij}(\beta) = \exp(X_{ij}\beta) / \sum_k \exp(X_{ik}\beta)$. *Conditional on a
draw* $\beta$, the kernel $L_{ij}(\beta)$ is an ordinary logit and obeys IIA. It
is the **mixing** — integrating over $\beta$ — that breaks IIA in the
unconditional probabilities. This is the mirror image of the multinomial logit,
which mixes a logit kernel over *observed* heterogeneity; the mixed logit mixes
it over *unobserved* heterogeneity. Either way, flexible substitution comes from
averaging logit kernels across heterogeneity — the only question is whether that
heterogeneity is in the data or in the assumed distribution.

```{r setup}
library(choicer)
set_num_threads(2)
```

## Simulate correlated random coefficients

`simulate_mxl_data()` draws choices in which the coefficients on `w1` and `w2`
are themselves random and *correlated* across the population.

```{r sim}
sim <- simulate_mxl_data(N = 2000, J = 4, seed = 1)
sim
```

## Fit

A robust recipe for mixed logit: warm-start from a plain MNL, scale the
variables so the Hessian is well conditioned, and use enough Halton draws. Here
we estimate a full (correlated) covariance of the random coefficients.

```{r fit}
fit <- run_mxlogit(
  data            = sim$data,
  id_col          = "id",
  alt_col         = "alt",
  choice_col      = "choice",
  covariate_cols  = c("x1", "x2"),  # fixed coefficients
  random_var_cols = c("w1", "w2"),  # random coefficients
  rc_correlation  = TRUE,           # estimate their full covariance
  S               = 100L,           # Halton draws per person
  draws           = "generate",     # generate draws on the fly (low memory)
  seed            = 7L,
  scale_vars      = "sd",           # condition the Hessian across blocks
  se_method       = "bhhh"
)
summary(fit)
```

This vignette uses `se_method = "bhhh"` because the outer-product calculation is
fast and keeps package-build time short. For final empirical work, compare it
with the default analytical-Hessian standard errors; when the data come from a
choice-based or otherwise weighted sample, use `se_method = "sandwich"` so the
reported covariance is the robust WESML sandwich rather than the inverse weighted
Hessian.

> **Tip.** For real applications increase the number of draws (`S`) until your
> estimates are stable, and keep `scale_vars = "sd"`. If the solver struggles,
> pass an explicit `theta_init` (for example the MNL coefficients) and bounds on
> the Cholesky diagonal. See `inst/simulations/mxl_simulation.R` for a fully
> hardened example.

## Parameter recovery

```{r recovery}
recovery_table(fit, sim$true_params)
```

The `beta` rows are the fixed coefficients, the `sigma` rows describe the
covariance of the random coefficients (its Cholesky elements), and the `asc`
rows are the alternative-specific constants.

## Substitution is no longer share-proportional

Because tastes are heterogeneous, mixed logit breaks IIA: people who like one
alternative tend to like its close substitutes, so demand diverts to similar
alternatives rather than in proportion to shares. Diversion therefore depends on
*which* attribute is changing, so `diversion_ratios()` takes a `wrt_var`:

```{r diversion}
elasticities(fit, elast_var = "x2")
diversion_ratios(fit, wrt_var = "x2")
```

The rest of the toolkit — `predict()`, `wtp()`, `consumer_surplus()`, `blp()` —
behaves exactly as in the [getting-started vignette](choicer.html), integrating
over the distribution of tastes automatically.

## No free lunch: the flexibility lives in the tails

The mixed logit is a genuine generalization of the MNL — if tastes are in fact
homogeneous, the estimator simply returns a near-zero variance and you are back
to a logit. So the issue is not that the random coefficients *impose* substitution
patterns. The issue is **identification**: all of the model's extra flexibility
is carried by the mixing distribution $f(\beta)$, and $f(\beta)$ is hardest to
pin down exactly where it matters most for substitution and welfare — in the
**tails**.

Two consequences are worth keeping in front of you:

- **The tails drive the economics you report.** A lognormal price coefficient
  puts a slice of the population at near-zero price sensitivity, which can give a
  willingness-to-pay distribution with *no finite mean* and explosive welfare
  numbers. An unbounded normal coefficient implies a fraction of consumers with
  the *wrong sign* (who prefer paying more). These artifacts come from the
  assumed shape of $f$, not from the data, and the estimator will happily contort
  a tail to match an aggregate moment.

- **$f(\beta)$ is hard to identify and estimate in practice.** A single
  cross-section of choices — one decision per person, a fixed menu — carries
  little information about the spread of tastes. Reliable estimation typically
  needs **repeated choices from the same individual** (panel data), **substantial
  variation in choice sets or attributes across markets** (the BLP setting), or
  the rich, designed attribute variation of a stated-preference experiment.
  Without one of these, the random-coefficient variances are weakly identified
  and the estimates can be fragile.

Practical defenses follow directly: prefer **bounded or sign-constrained mixing
distributions** (triangular, censored normal) so a tail cannot run away;
**estimate in WTP-space** (Train & Weeks, 2005) to sidestep the ratio-of-normals
pathology; or use a **latent-class** specification when a handful of discrete
types is more credible than a continuous distribution. None of these is free —
each trades one assumption for another — but each spends the assumption where you
can defend it. The broader tradeoff is laid out in
[Choosing among logit models](choicer.html#choosing-among-logit-models).
