---
title: "Mixed logit and preference heterogeneity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed logit and preference heterogeneity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

The mixed (random-coefficients) logit lets tastes vary across people. Instead of
a single coefficient on each random attribute, choicer estimates a *distribution*
of coefficients, which relaxes the IIA property and allows realistic substitution
patterns. Estimation is by simulated maximum likelihood using Halton draws, with
the likelihood, gradient and Hessian evaluated in parallel C++.

```{r setup}
library(choicer)
set_num_threads(2)
```

## Simulate correlated random coefficients

`simulate_mxl_data()` draws choices in which the coefficients on `w1` and `w2`
are themselves random and *correlated* across the population.

```{r sim}
sim <- simulate_mxl_data(N = 2000, J = 4, seed = 1)
sim
```

## Fit

A robust recipe for mixed logit: warm-start from a plain MNL, scale the
variables so the Hessian is well conditioned, and use enough Halton draws. Here
we estimate a full (correlated) covariance of the random coefficients.

```{r fit}
fit <- run_mxlogit(
  data            = sim$data,
  id_col          = "id",
  alt_col         = "alt",
  choice_col      = "choice",
  covariate_cols  = c("x1", "x2"),  # fixed coefficients
  random_var_cols = c("w1", "w2"),  # random coefficients
  rc_correlation  = TRUE,           # estimate their full covariance
  S               = 100L,           # Halton draws per person
  draws           = "generate",     # generate draws on the fly (low memory)
  seed            = 7L,
  scale_vars      = "sd",           # condition the Hessian across blocks
  se_method       = "bhhh"
)
summary(fit)
```

> **Tip.** For real applications increase the number of draws (`S`) until your
> estimates are stable, and keep `scale_vars = "sd"`. If the solver struggles,
> pass an explicit `theta_init` (for example the MNL coefficients) and bounds on
> the Cholesky diagonal. See `inst/simulations/mxl_simulation.R` for a fully
> hardened example.

## Parameter recovery

```{r recovery}
recovery_table(fit, sim$true_params)
```

The `beta` rows are the fixed coefficients, the `sigma` rows describe the
covariance of the random coefficients (its Cholesky elements), and the `asc`
rows are the alternative-specific constants.

## Substitution is no longer share-proportional

Because tastes are heterogeneous, mixed logit breaks IIA: people who like one
alternative tend to like its close substitutes, so demand diverts to similar
alternatives rather than in proportion to shares. Diversion therefore depends on
*which* attribute is changing, so `diversion_ratios()` takes a `wrt_var`:

```{r diversion}
elasticities(fit, elast_var = "x2")
diversion_ratios(fit, wrt_var = "x2")
```

The rest of the toolkit — `predict()`, `wtp()`, `consumer_surplus()`, `blp()` —
behaves exactly as in the [getting-started vignette](choicer.html), integrating
over the distribution of tastes automatically.
