---
title: "Mixed logit and preference heterogeneity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed logit and preference heterogeneity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

The mixed (random-coefficients) logit lets tastes vary across decision units. Instead of
a single coefficient on each random attribute, choicer estimates a *distribution*
of coefficients. Substitution then reflects both observed covariates and the
estimated distribution of tastes. Estimation is by simulated maximum likelihood
using Halton draws, with the likelihood, gradient and Hessian evaluated in
parallel C++. By default `run_mxlogit()` estimates a **cross-sectional**
simulated likelihood: each choice situation's probability is integrated over
the taste distribution separately. When the same decision maker is observed
in several choice situations, `person_col` switches to the **panel**
likelihood instead (Revelt and Train 1998): one taste draw is held fixed
across that person's situations and their conditional probabilities are
integrated jointly, so repeated choices discipline the taste distribution
directly — see *Panel data: one taste draw per decision maker* below.

A useful way to see the mechanism is to write the mixed-logit probability as a
logit kernel averaged over the taste distribution,
$P_{ij} = \int L_{ij}(\beta)\, f(\beta)\, d\beta$, where
$L_{ij}(\beta) = \exp(X_{ij}\beta) / \sum_k \exp(X_{ik}\beta)$. *Conditional on a
draw* $\beta$, the kernel is an ordinary logit. The empirical content of the
mixed logit is in the averaging: people with different $\beta$'s place different
values on the same attributes, so demand leaving an alternative need not go to
the same destinations for all consumers. For counterfactual work, the distinction
between observed heterogeneity and the estimated mixing distribution is central:
flexibility that is not disciplined by the available variation is supplied by the
maintained distribution $f(\beta)$.

```{r setup}
library(choicer)
set_num_threads(2)
```

## Simulate correlated random coefficients

`simulate_mxl_data()` draws choices in which the coefficients on `w1` and `w2`
are themselves random and *correlated* across the population.

```{r sim}
sim <- simulate_mxl_data(N = 2000, J = 4, seed = 1)
sim
```

## Fit

A robust recipe for mixed logit: for a large model, warm-start from a plain MNL
(as `inst/simulations/mxl_simulation.R` does); let the optimizer work in
coordinates scaled by the curvature at the start values (the default,
`scale_vars = "bhhh"`); and use enough Halton draws. Here we estimate a full
(correlated) covariance of the random coefficients.

```{r fit}
fit <- run_mxlogit(
  data            = sim$data,
  id_col          = "id",
  alt_col         = "alt",
  choice_col      = "choice",
  covariate_cols  = c("x1", "x2"),  # fixed coefficients
  random_var_cols = c("w1", "w2"),  # random coefficients
  rc_correlation  = TRUE,           # estimate their full covariance
  S               = 100L,           # Halton draws per choice situation (no person_col here)
  draws           = "generate",     # generate draws on the fly (low memory)
  seed            = 7L,
  scale_vars      = "bhhh",         # the optimizer's coordinates (the default)
  se_method       = "bhhh"
)
summary(fit)
```

This vignette uses `se_method = "bhhh"` because the outer-product calculation is
fast and keeps package-build time short. For final empirical work, compare it
with the default analytical-Hessian standard errors; when the data come from a
choice-based or otherwise weighted sample, use `se_method = "sandwich"` so the
reported covariance is the robust WESML sandwich rather than the inverse weighted
Hessian. If WESML observations are also dependent within people or markets, use
the weighted cluster sandwich (`se_method = "cluster"`, with `cluster_col`) and
have enough independent clusters for its asymptotics.

> **Tip.** For real applications increase the number of draws (`S`) until your
> estimates are stable, and keep the default `scale_vars = "bhhh"`. Stability in `S` is a
> statistical requirement, not just numerical hygiene: simulated ML with a
> fixed number of draws is biased — the log of an unbiased probability
> simulator is not unbiased — and the classical asymptotics require `S` to
> grow with the sample (see the
> [math companion](https://fpcordeiro.github.io/choicer/articles/mixed_logit_math.html) for the conditions). If the solver
> struggles, pass an explicit `theta_init`: place the MNL slope and ASC estimates
> in the corresponding MXL parameter blocks and initialize the additional mean
> and Cholesky coordinates deliberately. Bounds can keep Cholesky diagonals away
> from numerically pathological regions. See
> `inst/simulations/mxl_simulation.R` for a fully hardened example.

## Parameter recovery

```{r recovery}
recovery_table(fit, sim$true_params)
```

The `beta` rows are the fixed coefficients, the `sigma` rows describe the
covariance of the random coefficients (its Cholesky elements), and the `asc`
rows are the alternative-specific constants.

## Substitution through taste heterogeneity

With random coefficients, diversion is mediated by the distribution of tastes. If
the estimated mixing distribution captures economically meaningful heterogeneity,
people who value one alternative tend to value nearby substitutes on that latent
margin. Diversion can therefore depend on *which* attribute is changing, so
`diversion_ratios()` takes a `wrt_var`:

```{r diversion}
elasticities(fit, elast_var = "x2")
diversion_ratios(fit, wrt_var = "x2")

# For a random-coefficient attribute the perturbation coordinate matters.
elasticities(fit, elast_var = "w2", is_random_coef = TRUE)
diversion_ratios(fit, wrt_var = "w2", is_random_coef = TRUE)
```

The rest of the toolkit — `predict()`, `wtp()`, `consumer_surplus()`, `blp()` —
uses the same fitted object as in the [getting-started vignette](choicer.html).
Prediction and share inversion integrate over the fitted taste distribution;
`wtp()` respects choicer's supported coefficient parameterizations. In v0.2.0,
`consumer_surplus()` requires a fixed price coefficient because inverse moments
of a random price coefficient need not exist.

## Identification and tails

The mixed logit is a genuine generalization of the MNL — if tastes are in fact
homogeneous, the estimator simply returns a near-zero variance and you are back
to a logit. The issue is not that random coefficients are intrinsically fragile.
The issue is identification: the additional substitution structure is carried by
the mixing distribution $f(\beta)$, and $f(\beta)$ is often hardest to pin down
where it matters most for welfare and diversion — in the tails.

Two consequences are worth keeping in front of you:

- **The tails drive the economics you report.** A lognormal price coefficient
  puts a slice of the population at near-zero price sensitivity, which can give a
  willingness-to-pay distribution with *no finite mean* and explosive welfare
  numbers. An unbounded normal coefficient implies a fraction of consumers with
  the *wrong sign* (who prefer paying more). These artifacts come from the
  assumed shape of $f$, not from the data, and the estimator will happily contort
  a tail to match an aggregate moment.

- **$f(\beta)$ is hard to identify and estimate in practice.** A single
  cross-section of choices — one decision per person, a fixed menu — carries
  little information about the spread of tastes. In mixed-logit research,
  reliable estimation typically uses **repeated choices from the same individual**
  (panel data), **substantial
  variation in choice sets or attributes across markets** (the BLP setting), or
  the rich, designed attribute variation of a stated-preference experiment.
  Without one of these, the random-coefficient variances are weakly identified
  and the estimates can be fragile. `run_mxlogit()` exploits the first source
  when `person_col` is supplied: one taste draw is then held fixed across a
  person's choice situations, so it is their *joint* probability that gets
  integrated over $f(\beta)$ (Revelt and Train 1998) — see *Panel data: one
  taste draw per decision maker* below. Without `person_col`, the default
  cross-sectional likelihood integrates each situation separately and does not
  use that within-person link; cluster-robust standard errors (`cluster_col=`,
  or `vcov(fit, type = "cluster")`; see
  [Which standard errors, and when](inference.html)) then repair the
  *inference* for within-person dependence without changing what the point
  estimates target. The hierarchical Bayesian logit,
  [`run_hmnlogit()`](hb.html), is the Bayesian alternative to the panel
  likelihood: the same taste-sharing idea, plus partial pooling of
  alternative-level effects and posterior rather than asymptotic inference.

## Panel data: one taste draw per decision maker

**Estimand.** For decision maker $n$ observed in choice situations
$t = 1, \dots, T_n$, the panel likelihood holds one taste draw
$\beta_n \sim f(\beta \mid \theta)$ fixed across all of $n$'s situations and
integrates their *joint* probability (Revelt and Train 1998):
$$L_n(\theta) = \int \prod_{t=1}^{T_n} P_{nt}(j_{nt} \mid \beta)\, f(\beta \mid \theta)\, d\beta.$$
The target is the same taste distribution $f(\beta \mid \theta)$; what
differs is the maintained model of a person's repeated choices. The default
likelihood integrates every situation on its own — the full likelihood when
tastes are redrawn each situation or each person is observed once.

**Why repeated choices identify $\Sigma$.** A person who keeps picking the
same kind of alternative, situation after situation, is revealing that their
tastes sit away from the population average and stay there — that
persistence is exactly the variation that identifies $\Sigma$ most sharply
(the marginals identify it too, but only through the shape of the mixture).
The
cross-sectional likelihood cannot see it: integrating each situation on its
own treats a person's five consistent choices the same as five different
people who each chose once. It still targets the same population mixing
distribution — situation by situation, each probability is correctly
specified, so it stays consistent (a composite, or pseudo-, likelihood) — but
it is statistically inefficient, and the loss falls squarely on $\Sigma$.

**Fitting the panel likelihood.** Point `person_col` at the respondent
identifier; `id_col` still identifies choice situations, and must stay unique
across the whole data set (situation ids do not restart within person). `S`
becomes draws per decision maker rather than per choice situation, since one
draw block is now shared across all of a person's situations. A product of
$T_n$ per-situation probabilities is more sharply peaked in $\beta$ than a
single choice probability and needs more draws to simulate accurately — the
recovery script in `inst/simulations/mxl_panel_simulation.R` uses `S = 1000`
at `T = 10`; this vignette keeps `S = 100` for speed.

```{r sim-panel}
panel_sim <- simulate_mxl_data(N = 300, T = 6, J = 4, seed = 1)
```

`T = 6` gives each of the 300 decision makers six choice situations, sharing
one realized taste vector (`panel_sim$true_params$gamma_i[, n]`) across them.

```{r fit-panel}
panel_fit <- run_mxlogit(
  data            = panel_sim$data,
  id_col          = "id",
  alt_col         = "alt",
  choice_col      = "choice",
  covariate_cols  = c("x1", "x2"),
  random_var_cols = c("w1", "w2"),
  rc_correlation  = TRUE,
  person_col      = "pid",          # one taste draw per pid, shared across its T tasks
  S               = 100L,           # Halton draws per decision maker
  draws           = "generate",
  seed            = 7L
)
summary(panel_fit)
```

```{r recovery-panel}
recovery_table(panel_fit, panel_sim$true_params)
```

**What changes in inference.** With `person_col`, scores are already per
decision maker, so `se_method = "sandwich"` (equivalently
`vcov(panel_fit, type = "robust")`) is robust to within-person dependence out
of the box — there is nothing left within a person to cluster on.
`cluster_col` is for coarser groups that nest decision makers (households,
markets); it must be constant within each person. Weights, if supplied, are
decision-maker weights, constant within person, since the objective is
$\sum_n w_n \log L_n$; WESML (choice-based) weights are not supported with
`person_col`, because a choice-based weight varies with the alternative a
person chose, which can differ across their situations. `nobs()` still counts
choice situations; a panel fit additionally reports `n_persons`, printed as
"Respondents" above.

As a check on the efficiency argument above: refitting the same data
cross-sectionally, clustering on `pid` for valid inference, repairs the
standard errors' *validity* but not their size.

```{r cs-contrast}
cs_fit <- run_mxlogit(
  data            = panel_sim$data,
  id_col          = "id",
  alt_col         = "alt",
  choice_col      = "choice",
  covariate_cols  = c("x1", "x2"),
  random_var_cols = c("w1", "w2"),
  rc_correlation  = TRUE,
  cluster_col     = "pid",          # inference fix only: still one draw per situation
  S               = 100L,
  draws           = "generate",
  seed            = 7L
)
se_ratio <- sqrt(diag(vcov(cs_fit, type = "cluster"))) / sqrt(diag(vcov(panel_fit)))
round(se_ratio, 2)
```

The ratio sits close to one on the mean-utility parameters (`beta`, ASCs —
about 1.0-1.2 here) but visibly above one on the Cholesky block — the cost of
never sharing a taste draw within a person, exactly where a composite
likelihood's efficiency loss concentrates. (`cs_fit`'s numerator is the
cluster-robust sandwich; `panel_fit`'s denominator is the default
analytical-Hessian SE, so the ratio also reflects that choice of estimator,
not only the likelihood.) Because it targets a different (composite)
likelihood, the cross-sectional fit's log-likelihood, AIC and BIC are not
comparable to the panel fit's.

**Individual-level tastes.** `conditional_tastes()` computes, for each
decision maker, the mean and SD of $\beta_n$ conditional on the choices they
made (Revelt and Train 2000; Train 2009, ch. 11) — the frequentist analogue of
the hierarchical Bayesian `beta_i` summaries of [`run_hmnlogit()`](hb.html):

```{r conditional-tastes}
ct <- conditional_tastes(panel_fit)
ct
```

```{r cs-conditional-tastes}
cs_ct <- conditional_tastes(cs_fit)
```

Read this carefully: `ct$mean[, n]` is *not* decision maker $n$'s own
$\beta_n$. It is the average taste among people who would make $n$'s choices,
so it shrinks toward the population mean — more so for people with fewer
situations — and its spread across people understates the population spread
in $\Sigma$. The printed `revealed` and `residual` columns decompose
population variance by the law of total variance,
$\mathrm{Var}(\beta) = \mathrm{Var}(E[\beta \mid y]) + E[\mathrm{Var}(\beta \mid y)]$;
they should sum to about one. This is a numerical check, not a specification
test: the random-coefficient covariance $\Sigma$ is always freely estimated,
so this identity is the likelihood's first-order condition for $\Sigma$ and
holds at a converged fit whatever the true taste distribution — a departure
points to too few draws `S` or an unconverged fit. (Here `rc_mean = FALSE`
fixes the random-coefficient means at zero rather than estimating them
freely; a `mean_cond` far from zero below would flag that restriction.) In
this example the conditional means correlate
`r round(cor(ct$mean["w1", ], panel_sim$true_params$gamma_i["w1", colnames(ct$mean)]), 2)`
(`w1`) and
`r round(cor(ct$mean["w2", ], panel_sim$true_params$gamma_i["w2", colnames(ct$mean)]), 2)`
(`w2`) with the realized tastes in `panel_sim$true_params$gamma_i`. At the
true $\theta$, shrinkage sets this correlation exactly:
$\mathrm{cor}(E[\beta \mid y], \beta) = sd(E[\beta \mid y]) / sd(\beta) =
\sqrt{\texttt{revealed}}$, so the same shrinkage that keeps `revealed` below
one keeps the correlation below one too. A single choice reveals little about
a person's tastes, so fitting without `person_col` leaves `revealed` much
smaller —
`r round(cs_ct$population["w1", "revealed"], 2)` vs.
`r round(ct$population["w1", "revealed"], 2)` for `w1`,
`r round(cs_ct$population["w2", "revealed"], 2)` vs.
`r round(ct$population["w2", "revealed"], 2)` for `w2` — than with repeated
choices; conditional means only earn their keep with repeated choices.

Using conditional means as regressors in a second-stage analysis is worse
than merely ignoring their sampling and simulation error: the draw weights
behind each conditional mean are not exported, so regressing on `ct$mean`
directly measures tastes with error and biases the second-stage coefficient
toward zero (attenuation), on top of understating its standard error. Prefer
entering the observables in utility instead — interactions with the
random-coefficient columns, or `rc_mean` shifts — or, if a second-stage
regression is unavoidable, propagate the uncertainty with a
decision-maker-level bootstrap rather than treating conditional means as if
they were data.

## A research checklist

For a reportable `run_mxlogit()` application:

1. Fit a transparent MNL benchmark and use its estimates to inform explicit
   `theta_init` values. The default start moves with the covariates' units (a
   normal random coefficient's standard deviation starts at one over its
   covariate's typical step: 1 for a dummy, the standard deviation for a
   dense covariate), so the optimizer's path barely depends on them (exactly
   so, in exact arithmetic, under `scale_vars = "sd"`); a shifted log-normal
   coefficient's random part, though, has median one in its covariate's units:
   put such a covariate in units where the coefficient is of order one.
2. Run several manually supplied starting values and retain their objective,
   convergence code and substantive outputs. choicer does not yet automate
   multistart estimation.
3. Increase `S` until coefficients, covariance elements, standard errors, WTP,
   elasticities and diversion ratios are stable—not merely until the optimizer
   reports convergence.
4. Compare choicer's supported normal and shifted-lognormal restrictions where
   economically meaningful, and show what their tails imply for signs and WTP.
5. Compare the analytical-Hessian and BHHH results; use robust or WESML inference
   for weighting/misspecification and cluster inference for repeated sampling
   units as appropriate.
6. On panel data, decide what you can defend about tastes within a person
   before choosing the standard error: fit `person_col` when tastes can be
   defended as constant within a decision maker's situations, and keep the
   clustered cross-sectional fit when they cannot — it relies only on each
   situation's marginal and pays in precision, mostly on $\Sigma$. Add
   `run_hmnlogit()` when partially-pooled alternative effects or posterior
   uncertainty are wanted too.
7. Compare substitution and welfare with the MNL or NL benchmark and state which
   data variation, rather than the mixing distribution alone, earns the change.

Bounded/censored mixing distributions, WTP-space estimation, and latent-class
logit can be useful alternative modeling strategies, but they are **not
implemented in choicer v0.2.0**. They require another implementation today and
are roadmap candidates here; they should not be presented as options to
`run_mxlogit()`. The broader tradeoff is laid out in [Choosing among choice
models](choicer.html#choosing-among-choice-models).

## References

McFadden, D. and Train, K. (2000). Mixed MNL models for discrete response.
*Journal of Applied Econometrics*, 15(5), 447-470.

Revelt, D. and Train, K. (1998). Mixed logit with repeated choices:
households' choices of appliance efficiency level. *Review of Economics and
Statistics*, 80(4), 647-657.

Revelt, D. and Train, K. (2000). Customer-specific taste parameters and mixed
logit: households' choice of electricity supplier. Working Paper E00-274,
Department of Economics, University of California, Berkeley.

Train, K. E. (2009). *Discrete Choice Methods with Simulation* (2nd ed.).
Cambridge University Press.

Train, K. and Weeks, M. (2005). Discrete choice models in preference space
and willingness-to-pay space. In R. Scarpa and A. Alberini (Eds.),
*Applications of Simulation Methods in Environmental and Resource Economics*.
Springer.
