---
title: "Nested logit and relaxed IIA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nested logit and relaxed IIA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

Nested logit groups alternatives into *nests* of close substitutes. Within a
nest, alternatives share an unobserved component, so substitution is stronger
inside a nest than across nests — the independence-of-irrelevant-alternatives
(IIA) property holds within nests but not between them. Each nest has a
dissimilarity parameter λ in (0, 1\]: smaller λ means tighter substitution
inside the nest.

```{r setup}
library(choicer)
set_num_threads(2)
```

## Simulate a nested process

`simulate_nl_data()` builds two nests of inside goods plus an outside option,
with known dissimilarity parameters.

```{r sim}
sim <- simulate_nl_data(N = 4000, seed = 1)
sim
```

## Fit

Supply the nest membership column; choicer estimates the coefficients, the
alternative-specific constants and the nest dissimilarity parameters jointly,
using an analytical gradient and Hessian.

```{r fit}
fit <- run_nestlogit(
  data                   = sim$data,
  id_col                 = "id",
  alt_col                = "j",
  choice_col             = "choice",
  covariate_cols         = c("X", "W"),
  nest_col               = "nest",
  use_asc                = TRUE,
  include_outside_option = TRUE,
  outside_opt_label      = 0L
)
summary(fit)
```

## Parameter recovery

The `lambda` rows are the nest dissimilarity parameters — the part that is
unique to nested logit.

```{r recovery}
recovery_table(fit, sim$true_params)
```

## Nest-consistent elasticities

This is the payoff of nesting: cross-elasticities are larger for alternatives in
the *same* nest than for alternatives in different nests. choicer's
`elasticities()` respects the nest structure automatically.

```{r elast}
elasticities(fit, elast_var = "W")
diversion_ratios(fit)
```

## Share inversion with BLP

`blp()` runs the Berry-Levinsohn-Pakes contraction to recover the mean utilities
that reproduce a set of target shares — useful for calibration and demand
estimation from aggregate data. For strongly-nested models a damping factor
below 1 stabilizes the iteration.

```{r blp}
target_shares <- predict(fit, type = "shares")
head(blp(fit, target_shares, damping = 0.5))
```

As always, `predict()`, `wtp()` and `consumer_surplus()` are available on the
fitted object with the same syntax used throughout choicer.
