---
title: "Nested logit and relaxed IIA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nested logit and relaxed IIA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

Nested logit groups alternatives into *nests* of close substitutes. Within a
nest, alternatives share an unobserved component, so substitution is stronger
inside a nest than across nests — the independence-of-irrelevant-alternatives
(IIA) property holds within nests but not between them. Each nest has a
dissimilarity parameter λ in (0, 1\]: smaller λ means tighter substitution
inside the nest.

Think of nested logit as the **parsimonious middle ground** between the
multinomial and mixed logits. It buys a controlled departure from IIA — genuine
within-nest correlation in the unobserved utility — at the cost of just one extra
parameter per nest, and it stays globally well-behaved and cheap to estimate (a
closed-form GEV likelihood, no simulation). The price you pay is a strong prior:
*you* must specify the nesting tree in advance, and the model only permits
correlation within the nests you draw. When the right grouping is obvious from
the application (travel modes, product categories), that is a small price; when
it is not, the result can be sensitive to how you nest. See
[Choosing among logit models](choicer.html#choosing-among-logit-models) for how
this tradeoff compares with the alternatives.

```{r setup}
library(choicer)
set_num_threads(2)
```

## Simulate a nested process

`simulate_nl_data()` builds two nests of inside goods plus an outside option,
with known dissimilarity parameters.

```{r sim}
sim <- simulate_nl_data(N = 4000, seed = 1)
sim
```

## Fit

Supply the nest membership column; choicer estimates the coefficients, the
alternative-specific constants and the nest dissimilarity parameters jointly,
using an analytical gradient and Hessian.

```{r fit}
fit <- run_nestlogit(
  data                   = sim$data,
  id_col                 = "id",
  alt_col                = "j",
  choice_col             = "choice",
  covariate_cols         = c("X", "W"),
  nest_col               = "nest",
  use_asc                = TRUE,
  include_outside_option = TRUE,
  outside_opt_label      = 0L
)
summary(fit)
```

## Parameter recovery

The `lambda` rows are the nest dissimilarity parameters — the part that is
unique to nested logit.

```{r recovery}
recovery_table(fit, sim$true_params)
```

## Nest-consistent elasticities

This is the payoff of nesting: cross-elasticities are larger for alternatives in
the *same* nest than for alternatives in different nests. choicer's
`elasticities()` respects the nest structure automatically.

```{r elast}
elasticities(fit, elast_var = "W")
diversion_ratios(fit)
```

## Share inversion with BLP

`blp()` runs the Berry-Levinsohn-Pakes contraction to recover the mean utilities
that reproduce a set of target shares — useful for calibration and demand
estimation from aggregate data. For strongly-nested models a damping factor
below 1 stabilizes the iteration.

```{r blp}
target_shares <- predict(fit, type = "shares")
head(blp(fit, target_shares, damping = 0.5))
```

As always, `predict()`, `wtp()` and `consumer_surplus()` are available on the
fitted object with the same syntax used throughout choicer.
