---
title: "Choice-based sampling and WESML weights"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choice-based sampling and WESML weights}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

Choice data are often sampled by outcome. A hospital-choice researcher may
oversample rare hospitals; a school-choice researcher may sample the same number
of students from each enrolled school; a stated-preference analyst may balance
observed choices by design. In these data the sample choice shares are not the
population choice shares. Treating the sample as random therefore changes the
likelihood target.

Manski and Lerman's (1977) weighted exogenous sample maximum likelihood (WESML)
correction weights each choice situation by

$$
w_i = \frac{Q_{j(i)}}{H_{j(i)}},
$$

where $j(i)$ is the alternative chosen by situation $i$, $Q_j$ is the population
share choosing alternative $j$, and $H_j$ is the corresponding sample share.
choicer provides two helpers:

- `sample_by_choice()` draws a choice-based sample from a population frame and
  attaches WESML weights.
- `wesml_weights()` computes the same weights when you already have a sample and
  know the population shares `Q`.

```{r setup}
library(choicer)
library(data.table)
set_num_threads(2)
```

## Build a population

For exposition, start from a simulated population. In empirical work, the
population shares `Q` usually come from administrative totals, market shares, or
survey weights external to the choice-based estimation sample.

```{r population}
sim <- simulate_mnl_data(
  N = 4000,
  J = 4,
  seed = 11,
  outside_option = FALSE,
  vary_choice_set = FALSE
)

pop <- as.data.table(sim$data)
Q <- prop.table(table(pop[choice == 1, alt]))
round(Q, 3)
```

## Draw a choice-based sample

Now sample the same number of choice situations from each chosen alternative.
This keeps whole choice situations together: if an id is sampled, all of its
alternative rows are retained.

```{r sample}
cb <- sample_by_choice(
  pop,
  id_col = "id",
  alt_col = "alt",
  choice_col = "choice",
  n_per_alt = 400L,
  seed = 12L
)

strata <- sort(names(attr(cb, "Q")))
rbind(
  population = attr(cb, "Q")[strata],
  sample     = attr(cb, "H")[strata]
) |> round(3)

cb[choice == 1, .(id, chosen_alt = alt, .wesml_weight)][1:8]
```

The sample choice shares are deliberately equalized, but the attached weights
restore the population shares in the weighted likelihood. The weights are
constant within id and repeated across that id's alternative rows.

## Weighted point estimates

The difference between an unweighted fit and a WESML fit is easiest to see with a
fast MNL example. The population fit is not available in ordinary applications;
it is included here only as a benchmark because this vignette simulated the full
population.

```{r fit-mnl}
fit_pop <- run_mnlogit(
  data = pop,
  id_col = "id",
  alt_col = "alt",
  choice_col = "choice",
  covariate_cols = c("x1", "x2")
)

fit_unweighted <- run_mnlogit(
  data = cb,
  id_col = "id",
  alt_col = "alt",
  choice_col = "choice",
  covariate_cols = c("x1", "x2")
)

id_weights <- unique(cb[, .(id, .wesml_weight)])[order(id)]

fit_wesml <- run_mnlogit(
  data = cb,
  id_col = "id",
  alt_col = "alt",
  choice_col = "choice",
  covariate_cols = c("x1", "x2"),
  weights = id_weights$.wesml_weight
)

coef_compare <- cbind(
  population_sample = coef(fit_pop),
  choice_sample_unweighted = coef(fit_unweighted),
  choice_sample_wesml = coef(fit_wesml)
)

round(coef_compare, 3)

share_compare <- rbind(
  population_sample = drop(predict(fit_pop, type = "shares")),
  choice_sample_unweighted = drop(predict(fit_unweighted, type = "shares")),
  choice_sample_wesml = drop(predict(fit_wesml, type = "shares"))
)
colnames(share_compare) <- as.character(fit_pop$alt_mapping[[2]])
round(share_compare, 3)
```

The unweighted estimator treats the equalized sample shares as population
shares. WESML reweights the sampled choice situations back to the population
choice shares; with ASCs in the model, the correction is especially visible in
the alternative-specific constants and fitted shares. In a single finite sample
the WESML estimates need not be closer to the population benchmark parameter by
parameter, but they target the population likelihood under the choice-based
sampling design.

`run_mnlogit()` accepts WESML weights through the `weights` argument. The example
above sorts the id-level weights because `prepare_mnl_data()` sorts choice
situations by id before estimation.

## Mixed logit inference

For mixed logit, pass the row-level weight column directly and request the
robust sandwich covariance:

```{r mxl-template, eval = FALSE}
fit_mxl <- run_mxlogit(
  data = cb,
  id_col = "id",
  alt_col = "alt",
  choice_col = "choice",
  covariate_cols = "x1",
  random_var_cols = "x2",
  weights_col = ".wesml_weight",
  se_method = "sandwich",
  S = 200L
)
```

The distinction matters. With non-uniform WESML weights, the inverse weighted
Hessian is not the appropriate covariance estimator. The sandwich uses the
weighted Hessian as bread and the squared-weight outer product of scores as
meat; it is invariant to arbitrary rescaling of the weights.

If the data came from `sample_by_choice()`, choicer records a
`choice_sampling` attribute and the weight column name. The explicit
`weights_col` argument is still recommended in scripts because it makes the
estimation target visible.

## Starting from an existing sample

When the choice-based sample already exists, provide the population shares `Q`
directly:

```{r attach}
cb2 <- copy(cb)
cb2[, .wesml_weight := NULL]

cb2 <- wesml_weights(
  cb2,
  id_col = "id",
  alt_col = "alt",
  choice_col = "choice",
  Q = attr(cb, "Q"),
  attach = TRUE
)

attr(cb2, "choice_sampling")
```

The names of `Q` must match the chosen-alternative strata exactly after coercion
to character. This strict matching is intentional: silently dropping a realized
stratum would change the target population.

## References

Manski, C. F. and Lerman, S. R. (1977). The estimation of choice probabilities
from choice based samples. *Econometrica*, 45(8), 1977-1988.

Train, K. E. (2009). *Discrete Choice Methods with Simulation* (2nd ed.).
Cambridge University Press, Section 3.7.
